---
title: "PM 2.5 Analysis"
output: html_notebook
---

```{r}
rm(list=ls())
```


```{r}
library(tidyverse)
library(stringr)
library(lubridate)
```




```{r}
## download and unzip data to working directory
df1 <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2FNEI_data.zip"
download.file(df1, "PM25.zip")
unzip("PM25.zip", exdir = ".")
```
```{r}
## load data
NEI <- readRDS("summarySCC_PM25.rds")
SCC <- readRDS("Source_Classification_Code.rds")
```

```{r}
## Review data
head(NEI)
head(SCC)
```

1. Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.

```{r}
## rename Year and change type to factor
NEI <- NEI %>% rename(Year = year)
NEI <- NEI %>% mutate(Year=factor(Year))


```


```{r}
head(NEI)
head(SCC)
```
```{r}
## summarize emissions by year and divided it by 1000
NEITotal <- group_by(NEI,Year) %>% summarize(Emissions = sum(Emissions, na.rm=TRUE)) %>% mutate(Emissions=Emissions/1000)
head(NEITotal)
```

1. Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.
```{r}
barplot(NEITotal$Emissions, names = NEITotal$Year, xlab = "Year", ylab = expression("Emissions of PM"[25]* " (in kilotons)"), main = expression("Total emissions from PM"[25]* " in US (1999-2008)"), col = c("red", "yellow", "blue", "green"))
dev.copy(png, file ="plot1.png")
dev.off()
```
2. Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == “24510”) from 1999 to 2008? Use the base plotting system to make a plot answering this question.

Answer: 
```{r}
## Filtered fips = Baltimore City and summarized it by emissions ~ year in kilotons
baltimorefilter <- NEI %>% filter(fips=="24510")
   
BaltimoreTotal <- baltimorefilter %>% group_by(Year) %>% summarize(Emissions=sum(Emissions, na.rm = TRUE)/1000)

barplot(BaltimoreTotal$Emissions, names = BaltimoreTotal$Year, 
        xlab = "Year", ylab = expression("Emissions of PM"[25]* " 
                                         (in kilotons)"), 
        main = expression("Baltimore total emissions from PM"[25]* 
                             " in US (1999-2008)"), col = c("red", "yellow", "blue", "green"))
dev.copy(png, file ="plot2.png")
dev.off()
```

3. Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999-2008 for Baltimore City? Which have seen increases in emissions from 1999-2008? Use the ggplot2 plotting system to make a plot answer this question.
```{r} 
## added type baltimore.  ggplot is divided via facet_wrap
BaltimoreTotal <- baltimorefilter %>% group_by(Year, type) %>% summarize(Emissions=sum(Emissions, na.rm = TRUE)/1000)

ggplot(data = BaltimoreTotal, aes(x=Year, y=Emissions,color=type))+geom_point()+facet_wrap(vars(type))+
   ggtitle(expression("Total Baltimore " ~ PM[2.5] ~ "Emissions by Type and Year")) +
      ylab(expression("Total Baltimore " ~ PM[2.5] ~ "Emissions (in kilotons)")) +
      xlab("Year") 
ggsave("plot3.png")
   
```
4. Across the United States, how have emissions from coal combustion-related sources changed from 1999-2008?
```{r}
## using stringr "str_detect". Reviewed "coal vs Coal" case sensitive and picked Coal.  
## coal produce 239 obs vs Coal produced 230 obs 

coalfil <- SCC %>% filter(str_detect(Short.Name, pattern = "Coal"))

coalNEI <- NEI %>% filter(SCC %in% coalfil$SCC)

coalTotal <- coalNEI %>% group_by(Year, type) %>% summarise(Emissions= sum(Emissions, na.rm = TRUE)/1000)

ggplot(data = coalTotal, aes(x=Year, y=Emissions,fill=type))+geom_col()+facet_wrap(vars(type))+
   ggtitle(expression("Total Coal " ~ PM[2.5] ~ "Emissions by Type and Year")) +
      ylab(expression("Total Coal " ~ PM[2.5] ~ "Emissions (in kilotons)")) +
      xlab("Year") 
ggsave("plot4.png")

```
5. How have emissions from motor vehicle sources changed from 1999-2008 in Baltimore City?
```{r}
baltfips <- NEI %>% select(fips:Year) %>% filter(fips=="24510" & type == "ON-ROAD")
BaltimoreVehicle <- baltfips %>% group_by(Year, type) %>% summarise(Emissions= sum(Emissions, na.rm = TRUE)/1000)

ggplot(data = BaltimoreVehicle, aes(x=Year, y=Emissions,fill=type))+geom_col()+facet_wrap(vars(type))+
   ggtitle(expression("Baltimore " ~ PM[2.5] ~ "Motor Vehicle Emissions by Year")) +
      ylab(expression("Baltimore " ~ PM[2.5] ~ "MV Emissions (in kilotons)")) +
      xlab("Year") 
ggsave("plot5.png")
```
6. Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == “06037”). Which city has seen greater changes over time in motor vehicle emissions?
```{r} 
Baltimore_LA <- c("24510", "06037")
BaltLAFil <- NEI %>% filter(fips==Baltimore_LA & type == "ON-ROAD")
BaltLAveh <- BaltLAFil %>% group_by(Year, fips, type) %>% summarise(Emissions= sum(Emissions, na.rm = TRUE)/1000)
BaltLAveh <- ungroup(BaltLAveh) %>% mutate(fips = recode(fips,`24510`="Baltimore City",  `06037`="Los Angeles City")) %>% rename(City=fips)



ggplot(data = BaltLAveh, aes(x=Year, y=Emissions, fill=City))+geom_col()+facet_wrap(vars(City))+
   ggtitle(expression("Baltimore and Los Angeles" ~ PM[2.5] ~ "Motor Vehicle Emissions by Year")) +
      labs(x = "Year", y = expression(~PM[2.5]~ "Motor Vehicle Emissions") ) +
      scale_colour_discrete(name = "City", labels = c("Los Angeles", "Baltimore")) +
      theme(legend.title = element_text(face = "bold"))
   
ggsave("plot5.png")
```


